Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-08-19 demystify.tex
updates

\renewcommand{\figurename}{Fig.}
\renewcommand{\tablename}{Tab.}
\def\Section{\S}
\renewcommand{\figureautorefname}{Fig.}
\renewcommand{\tableautorefname}{Tab.}
\makeatletter
\renewcommand{\chapterautorefname}{\S\@gobble}
\renewcommand{\sectionautorefname}{\S\@gobble}
\renewcommand{\subsectionautorefname}{\S\@gobble}
\renewcommand{\appendixautorefname}{\S\@gobble}
\makeatother

\newcommand{\mycomment}[3][\color{blue}]{{#1{{#2}: {#3}}}}
\newcommand{\tvn}[1]{\mycomment{TVN}{#1}}{}
\newcommand{\didi}[1]{\mycomment{Didier}{#1}}{}
\newcommand{\tl}[1]{\mycomment{ThanhLe}{#1}}{}
\newcommand{\red}[1]{{\color{red}{#1}}}
\newcommand{\xz}[1]{\mycomment{Xiaokuan}{[#1]}}{}

@@

\chapter{Summary}\label{sec:summary}

This section summarizes the main points of this guideline. It also gives you an overview to decide which specific topics you want to explore more thoroughly.

\begin{enumerate}
  \item Should you apply to a PhD CS program in the US?
        \begin{itemize}
          \item \emph{Yes, definitely}.  CS PhD study in the US is fully funded and admission into good universities is not any harder than in other countries (\autoref{sec:should}).
        \end{itemize}
  \item How is your application evaluated?
        \begin{itemize}
          \item Applications are evaluated by the \emph{PhD Admission} (adcom) committee and each application is typically reviewed by three faculty members (\autoref{sec:evalapps}).
          \item Individually faculty member \emph{cannot directly admit} a student---so do not email and ask if you have a chance. However, faculty can \emph{advocate} for a student and therefore increase their admission chance---so do contact and introduce yourself (\autoref{sec:contact}).
         \end{itemize}
  \item Application Materials
        \begin{itemize}
          \item The committee will look at various factors, but the most important ones are letters of recommendation (LORs),  statements of purpose (SOP), and research experience and potential, e.g., publications.
          \item \emph{LORs are very important}, but only if they are personalized and talk about your research ability (\autoref{sec:lor}).
          \item \emph{SOP is very important}. Write it in such a way that makes you \emph{stand out} (\autoref{sec:research-statement} and \autoref{sec:improve-your-chance})
          \item  \emph{Publications can greatly help}. Papers in good venues are concrete evidence that you have successfully engaged in research. But they are not required! (\autoref{sec:research-experience}).

          \item \emph{High grades probably won't help much} (unless you're from a well-known school), but bad ones likely will raise concerns (\autoref{sec:your-school}).

          \item \emph{Standard tests are not important} (\autoref{sec:standard-tests}). GRE typically \emph{is not} required. For standard English tests (not required for domestic students), just do enough to pass the minimum requirements.

          \item \emph{Contacting a prof. is recommended}, but do it \emph{properly} to get a reply (\autoref{sec:contact}).

          \item Getting an interview is typically a \emph{good sign}; but no interview does not mean rejection (\autoref{sec:interview}).

        \end{itemize}
  \item What to do after getting admitted?
        \begin{itemize}
                \item  Celebrating! Now it is your turn to evaluate the school!
          \item Attend \emph{Open House} to learn more about the place and \emph{interview} profs---they would be much more willing to talk to you now (\autoref{sec:accepted}).
        \end{itemize}
  \item Funding
        \begin{itemize}
          \item TAs, RAs, and fellowships are main funding sources for PhDs (\autoref{sec:funding}).  TAs (teaching assistantships) are provided by the department to help with classes (e.g., grading). RAs (research assistantships) are given by profs. to help with their research.  Fellowships, provided by the university, department, or external sources such as government or industry, give move flexibility but are very competitive.
        \end{itemize}
  \item Choosing School and Professors
        \begin{itemize}
          \item Not every professor, even those in CS, can advise or graduate CS PhD students (\autoref{sec:schoolsandprofs}).
        \end{itemize}

    \item Miscs and FAQS
        \begin{itemize}
          \item One reason why you were rejected (\autoref{sec:why-rejected}) is because you applied to AI/ML, a super competitive field in recent years with many applicants.

          \item Why rejection letters are sent so late? (\autoref{sec:late-rejection}). Grad programs often wait for the accepted students to make their decisions, typically by April 15, before sending out rejection letters.

          \item Increasing admission chance by being unique and standing out (\autoref{sec:improve-your-chance}).
          \item It takes about 10 minutes to review an application (\autoref{sec:ievaluate}).
          \item Sometimes adcom members discuss applicants, but in most cases they make independent decisions (\autoref{sec:adcom-discuss}).           
          \item Adcom members only review applications assigned to them (typically matching their expertise) and rarely get involved in other applications (\autoref{sec:read-all}).

          \item Being a minority can significantly help your application (\autoref{sec:urm}).
          \item You can successfully apply to CS PhD even if you have non-CS background (\autoref{sec:non-stem}).
          \item You do not need an MS (\autoref{sec:msrequirement}), and CS PhD takes about 5--7 years (\autoref{sec:time}). Compared to other countries, CS PhD in the US does not require an MS degree but has a longer PhD study time (\autoref{sec:non-us-differences}).
          \item What should you call your profs.? (\autoref{sec:address}) Many possibilities (e.g., prof., Dr., and even their firstnames), but probably not Mr. or Mrs.
          \item Your entire PhD program costs about \$400K in total, but you \emph{do not} pay for it (\autoref{sec:ra-cost})
          \item Ask your prof. if they can buy computer equipment and such for your research (\autoref{sec:buying-equipment}).
          %\item Despite some miserable stories on social media, many PhD students have good mentors, supportive lab mates, healthy working environment ... and are happy (\autoref{sec:happy}).
        \end{itemize}

    \item Appendices
    \begin{itemize}
      \item Specific advice for domestic students (\autoref{sec:domestic-students}).
      \item How to find research opportunities (\autoref{sec:research-opportunities}).
      \item Cultural differences between US and other countries (\autoref{sec:cultural}).
    \end{itemize}

\end{enumerate}

@@

@@

@@

After you submit your PhD application (e.g., in late Fall of year $X$ for the Fall admission of year $X+1$), it will first be screened
for general requirements, e.g., did you submit your transcripts and standard scores? did your reference writers submit their letters? Usually, this screening process is done through a central university system, i.e., not by CS faculty. If you pass this, your application will be forwarded to the CS department for further evaluation. If you don't,  the system will tell you what is missing and what you need to do. So pay attention to your email and check your application status regularly.

@@

Then, applications are reviewed by a \emph{PhD admission committee} (\textbf{adcom}) that consists of faculty members in CS\footnote{In some cases the committee can involve affiliated faculty from different disciplines.}. These adcom members have a wide range of expertise and background to ensure diverse perspectives in the evaluation process. The size and the review load of the adcom depend on the department size. At Mason, the PhD adcom typically has 15--20 faculty, and each committee member is assigned to review about 30 applications. Note that most large schools, including Mason, have separate adcoms for MS programs.

@@

@@

The focus of adcom is to evaluate your background and interest in research to see if you \textbf{fit into our PhD program}! The emphasis here is \emph{fitting}, which varies from school to school, faculty to faculty, and even from year to year.  However, in general, to evaluate your profile, we consider the following key indicators.

\section{Letters of Recommendation (LOR)}\label{sec:lor}

@@

\section{Statement of Purpose (SOP)}\label{sec:research-statement}

@@

@@

However, it is understandable that many students do not have the opportunity to publish papers. Thus, general writings, even those under submissions or rejected, would still be good (much better than nothing).  Be sure to upload your papers with your application and talk about them in your statement (see \autoref{sec:research-statement}).  Note that local conferences and non-English journals or conferences do
not carry as much weight since their quality is often unknown to US faculty. However, if you have published in such places, you should still upload them, mention them in your statement, and explain why they are good.

@@

\section{Your School and Grades}\label{sec:your-school}

@@

\section{Standard Tests (e.g., GRE and IELTS), Online Courses, and Certificates}\label{sec:standard-tests}

\subsection{GREs Do Not Matter Much} While a few schools still require taking the GRE (e.g., UCF), most good CS PhD programs in the US \textbf{no longer} require it. Thus, if you haven't taken the GRE or have bad scores, then don't waste time (re)taking it. Being optional really means optional, and not taking it will not hurt your application.

@@

\section{Interview and the Waiting Game}\label{sec:interview}

@@

@@

@@

For a full breakdown of how much a graduate student costs, see \autoref{sec:ra-cost}.

@@

@@

@@

@@

\section{Timeline: Why are rejection letters sent so late?}\label{sec:late-rejection}

@@

@@

\section{How long do you evaluate an application?}\label{sec:ievaluate}

@@

\section{Do adcom members talk to each other?}\label{sec:adcom-discuss}

@@

\section{Do adcom members read all applications?}\label{sec:read-all}

@@

\section{Member of an underrepresented or LGBTQ+ groups}\label{sec:urm}

@@

\section{My undergrad was not in CS or related areas}\label{sec:non-stem}

@@

\section{How do I call or address a professor?}\label{sec:address}

@@

\section{Buying Computer Equipment}\label{sec:buying-equipment}

@@

By contributing code, fixing bugs, implementing new features, or providing documentation, students can gain practical research experience and interact with experienced developers and researchers in the field. Not only you gain research experience, but you might be able to get a LoR from the project maintainer, and you should write about this experience in your SOP (\autoref{sec:research-statement}.)

@@

  \item[Adcom] Admission Committee. The group of faculty members who review applications and make admission decisions.
  \item [Adcom members] Faculty members who are part of the admission committee. These are the people who review applications and make admission decisions.
  \item[Adcom chair] The faculty member who leads the admission committee. This person is often not involved in individual admission decisions but oversees the process (e.g., resolve disputes, ensure fairness).
  \item[Advisor/Supervisor]: A faculty member who guides and mentors a PhD student throughout their research. This person typically plays a crucial role in a student's academic journey. In the US, the term ``adviser'' is more commonly used than ``supervisor''.
  \item[Cohort] A group of students who start a program at the same time and take classes together. You should get to know your cohort as they will be your colleagues and friends during your PhD.
  \item  [Deferred Admission] An option allowing admitted students to postpone their start date, typically by one year.
  \item [Diversity Statement] A document sometimes required in applications where applicants describe their contributions to diversity and inclusion, or how their background adds to the diversity of the academic community. This document is often required for faculty positions, but some universities also require it for graduate admissions.
  \item[Domestic Student] A student who is typically a US citizen or permanent resident (``greencard holder'').
        But sometimes it can also refer to students who did their undergrad (or MS) at a US university.
  \item[Field or Area] A specific research area, e.g., Machine Learning, Computer Vision, Software Engineering, etc.
  \item [In-state vs. Out-of-state] In-state tuition is the tuition rate for students who are residents of the state where the university is located. Out-of-state tuition is the tuition rate for students who are not residents of that state. In-state tuition is typically much lower than out-of-state tuition. PhD students typically do not have to pay tuition as it is covered by their funding, but this might be important for MS students.
  \item[LoR] Letter of Recommendation. A letter written by a professor or supervisor that assesses your qualifications and potential for graduate study (\autoref{sec:lor}).
  \item[SOP] Statement of Purpose. A document written by yourself to explains your research interests, background, and reasons for applying to a PhD program (\autoref{sec:research-statement}).

>>>>

\renewcommand{\figurename}{Fig.}
\renewcommand{\tablename}{Tab.}
\def\Section{\S}
\renewcommand{\figureautorefname}{Fig.}
\renewcommand{\tableautorefname}{Tab.}
\makeatletter
\renewcommand{\chapterautorefname}{\S\@gobble}
\renewcommand{\sectionautorefname}{\S\@gobble}
\renewcommand{\subsectionautorefname}{\S\@gobble}
\renewcommand{\appendixautorefname}{\S\@gobble}

\newcommand*\l@chapterinfo{\@nodottedtocline{0}{0.0em}{1.5em}}
\newcommand*\l@sectioninfo{\@nodottedtocline{1}{1.5em}{2.3em}}
\newcommand*\l@subsectioninfo{\@nodottedtocline{2}{3.8em}{3.2em}}
\newcommand*\l@subsubsectioninfo{\@nodottedtocline{3}{7.0em}{4.1em}}
\newcommand*\l@paragraphinfo{\@nodottedtocline{4}{10em}{5em}}
%\newcommand*\l@subparagraphinfo{\@nodottedtocline{5}{12em}{6em}}line{5}{12em}{6em}}

% Copied from the book class macro `\@dottedtocline`. Removed the dots and page number
\def\@nodottedtocline#1#2#3#4#5{%
  \ifnum #1>\c@tocdepth \else
    \vskip \z@ \@plus.2\p@
    {\leftskip #2\relax \rightskip \@tocrmarg \parfillskip -\rightskip
     \parindent #2\relax\@afterindenttrue
     \interlinepenalty\@M
     \leavevmode
     \@tempdima #3\relax
     \advance\leftskip \@tempdima \null\nobreak\hskip -\leftskip
     {#4}\nobreak
     \leaders\hbox{$\m@th
        \mkern \@dotsep mu\hbox{\,}\mkern \@dotsep
        mu$}\hfill
     \nobreak
     \hb@xt@\@pnumwidth{\hfil\normalfont \normalcolor }%
     \par}%
  \fi}

\makeatother

\def\chapterinfo#1{%
  \addcontentsline{toc}{chapterinfo}{%
    \noexpand\numberline{}\color{black}{#1}}%
}
\def\sectioninfo#1{%
  \addcontentsline{toc}{sectioninfo}{%
    \noexpand\numberline{}\color{black}{#1}}%
}
\def\subsectioninfo#1{%
    \addcontentsline{toc}{subsectioninfo}{%
    \noexpand\numberline{}\color{black}{#1}}%
}

\newcommand{\mycomment}[3][\color{blue}]{{#1{{#2}: {#3}}}}
\newcommand{\tvn}[1]{\mycomment{TVN}{#1}}{}
\newcommand{\didi}[1]{\mycomment{Didier}{#1}}{}
\newcommand{\tl}[1]{\mycomment{ThanhLe}{#1}}{}
\newcommand{\xz}[1]{\mycomment{Xiaokuan}{[#1]}}{}
\newcommand{\red}[1]{{\color{red}{#1}}}

@@

% \chapter{Summary}\label{sec:summary}

% This section summarizes the main points of this guideline. It also gives you an overview to decide which specific topics you want to explore more thoroughly.

  %\item Should you apply to a PhD CS program in the US?

          %\item \emph{Contacting a prof. is recommended}, but do it \emph{properly} to get a reply (\autoref{sec:contact}).

          %\item  (\autoref{sec:interview}).

    % \item Miscs and FAQS
    %     \begin{itemize}

    %       \item (\autoref{sec:read-all}).

    %       \item
    %       \item
    %       \item You do not need an MS (\autoref{sec:msrequirement}), and CS PhD takes about 5--7 years (\autoref{sec:time}). Compared to other countries, CS PhD in the US does not require an MS degree but has a longer PhD study time (\autoref{sec:non-us-differences}).
    %       \item
    %       \item
    %       \item
    %       %\item Despite some miserable stories on social media, many PhD students have good mentors, supportive lab mates, healthy working environment ... and are happy (\autoref{sec:happy}).
    %     \end{itemize}

    % \item Appendices
    % \begin{itemize}
    %   \item Specific advice for domestic students (\autoref{sec:domestic-students}).
    %   \item How to find research opportunities (\autoref{sec:research-opportunities}).
    %   \item Cultural differences between US and other countries (\autoref{sec:cultural}).
    % \end{itemize}

@@

\chapterinfo{\emph{Yes, definitely}.  CS PhD study in the US is fully funded and admission into good universities is not any harder than in other countries.}

@@

\chapterinfo{Applications are evaluated by the \emph{PhD Admission} (adcom) committee and each application is typically reviewed by three faculty members.}

@@

After you submit your PhD application (e.g., in late Fall of year $X$ for the Fall admission of year $X+1$), it will first be ``screened''
for general requirements, e.g., did you submit your transcripts and standard scores? did your reference writers submit their letters? Usually, this screening process is done through a central university system, i.e., not by CS faculty. If you pass this, your application will be forwarded to the CS department for further evaluation. If you don't,  the system will tell you what is missing and what you need to do. So pay attention to your email and check your application status regularly.

@@

\paragraph{Adcom} After screening applications are reviewed by a \emph{PhD admission committee} (\textbf{adcom}) that consists of faculty members in CS\footnote{In some cases the committee can involve affiliated faculty from different disciplines.}. These adcom members have a wide range of expertise and background to ensure diverse perspectives in the evaluation process. The size and the review load of the adcom depend on the department size. At Mason, the PhD adcom typically has 15--20 faculty, and each committee member is assigned to review about 30 applications. Note that most large schools, including Mason, have separate adcoms for MS programs.

@@

\chapterinfo{The committee will look at various factors, but the most important ones are letters of recommendation (LORs),  statements of purpose (SOP), and research experience and potential, e.g., publications.}

@@

The goal of adcom is to evaluate your research experience, potential, and interest to see if you \textbf{fit into our PhD program}! The emphasis here is \emph{fitting}, which varies from school to school, faculty to faculty, and even from year to year.  However, in general, to evaluate your profile, we consider the following key indicators.

\section{Letters of Recommendation (LOR)}\label{sec:lor}
\sectioninfo{\emph{LORs are very important}, but only if they are personalized and talk about your research ability.}

@@

\section{Statement of Purpose (SOP)}\label{sec:sop}
\sectioninfo{\\emph{SOP is very important}. Write it in such a way that makes you \emph{stand out}.}

@@

\sectioninfo{\emph{Publications can greatly help}. Papers in good venues are concrete evidence that you have successfully engaged in research. But they are not required!}

@@

However, it is understandable that many students do not have the opportunity to publish papers. Thus, general writings, even those under submissions or rejected, would still be good (much better than nothing).  Be sure to upload your papers with your application and talk about them in your statement (see \autoref{sec:sop}).  Note that local conferences and non-English journals or conferences do
not carry as much weight since their quality is often unknown to US faculty. However, if you have published in such places, you should still upload them, mention them in your statement, and explain why they are good.

@@

\section{Your School and Grades}\label{sec:your-school}
\sectioninfo{\emph{High grades probably won't help much} (unless you're from a well-known school), but bad ones likely will raise concerns.}

@@

\section{Standard Tests (e.g., GRE and IELTS), Online Courses, and Certificates}\label{sec:standard-tests}
\sectioninfo{\emph{Standard tests are not important}. GRE typically \emph{is not} required. For standard English tests (not required for domestic students), just do enough to pass the minimum requirements.}

\subsection{GREs Are Optional and Do Not Matter} While a few schools still require taking the GRE (e.g., UCF), most good CS PhD programs in the US \textbf{no longer} require it. Thus, if you haven't taken the GRE or have bad scores, then don't waste time (re)taking it. Being optional really means optional, and not taking it will not hurt your application.

@@

\section{Interview and the Waiting Game}\label{sec:interview}
\sectioninfo{Getting an interview is typically a \emph{good sign}; but no interview does not mean rejection}

@@

\chapterinfo{Celebrating! Now it is your turn to evaluate the school!
Attend \emph{Open House} to learn more about the place and \emph{interview} profs---they would be much more willing to talk to you now.}

@@

\chapterinfo{TAs, RAs, and fellowships are main funding sources for PhDs.  TAs (teaching assistantships) are provided by the department to help with classes (e.g., grading). RAs (research assistantships) are given by profs. to help with their research.  Fellowships, provided by the university, department, or external sources such as government or industry, give move flexibility but are very competitive.}

@@

For a full breakdown of how much a graduate student costs, see \autoref{sec:ra-cost}.
\sectioninfo{Your entire PhD program costs about \$400K in total, but you \emph{do not} pay for it.}

@@

\chapterinfo{Not every professor, even those in CS, can advise or graduate CS PhD students.}

@@

\subsectioninfo{Individually faculty member \emph{cannot directly admit} a student---so do not email and ask if you have a chance. However, faculty can \emph{advocate} for a student and therefore increase their admission chance---so do contact and introduce yourself.}

@@

\sectioninfo{One reason why you were rejected is because you applied to AI/ML, a super competitive field in recent years with many applicants.}

@@

\section{Timeline: Why are rejection letters sent so late?}\label{sec:late-rejection}
\sectioninfo{Why rejection letters are sent so late? Grad programs often wait for the accepted students to make their decisions, typically by April 15, before sending out rejection letters.}

@@

\sectioninfo{Increasing admission chance by being unique and standing out.}

@@

\section{How long do you evaluate an application?}\label{sec:ievaluate}
\sectioninfo{It takes about 10 minutes to review an application.}

@@

\section{Do adcom members talk to each other?}\label{sec:adcom-discuss}
\sectioninfo{Sometimes adcom members discuss applicants, but in most cases they make independent decisions.}

@@

\section{Do adcom members read all applications?}\label{sec:read-all}
\sectioninfo{Adcom members only review applications assigned to them (typically matching their expertise) and rarely get involved in other applications}

@@

\section{Member of an underrepresented or LGBTQ+ groups}\label{sec:urm}
\sectioninfo{Being a minority can significantly help your application.}

@@

\section{My undergrad was not in CS or related areas}\label{sec:non-stem}
\sectioninfo{You can successfully apply to CS PhD even if you have non-CS background.}

@@

\section{How do I call or address a professor?}\label{sec:address}
\sectioninfo{What should you call your profs.? Many possibilities (e.g., prof., Dr., and even their firstnames), but not Mr. or Mrs.}

@@

\section{Buying Computer Equipment}\label{sec:buying-equipment}
\sectioninfo{Ask your prof. if they can buy computer equipment and such for your research.}

@@

By contributing code, fixing bugs, implementing new features, or providing documentation, students can gain practical research experience and interact with experienced developers and researchers in the field. Not only you gain research experience, but you might be able to get a LoR from the project maintainer, and you should write about this experience in your SOP (\autoref{sec:sop}.)

@@

  \item[Adcom] Admission Committee. The group of faculty members who review applications and make admission decisions.
  \item [Adcom members] Faculty members who are part of the admission committee. These are the people who review applications and make admission decisions.
  \item[Adcom chair] The faculty member who leads the admission committee. This person is often not involved in individual admission decisions but oversees the process (e.g., resolve disputes, ensure fairness).
  \item[Advisor/Supervisor]: A faculty member who guides and mentors a PhD student throughout their research. This person typically plays a crucial role in a student's academic journey. In the US, the term ``adviser'' is more commonly used than ``supervisor''.
  \item[Cohort] A group of students who start a program at the same time and take classes together. You should get to know your cohort as they will be your colleagues and friends during your PhD.
  \item  [Deferred Admission] An option allowing admitted students to postpone their start date, typically by one year.
  \item [Diversity Statement] A document sometimes required in applications where applicants describe their contributions to diversity and inclusion, or how their background adds to the diversity of the academic community. This document is often required for faculty positions, but some universities also require it for graduate admissions.
  \item[Domestic Student] A student who is typically a US citizen or permanent resident (``greencard holder'').
        But sometimes it can also refer to students who did their undergrad (or MS) at a US university.
  \item[Field or Area] A specific research area, e.g., Machine Learning, Computer Vision, Software Engineering, etc.
  \item [In-state vs. Out-of-state] In-state tuition is the tuition rate for students who are residents of the state where the university is located. Out-of-state tuition is the tuition rate for students who are not residents of that state. In-state tuition is typically much lower than out-of-state tuition. PhD students typically do not have to pay tuition as it is covered by their funding, but this might be important for MS students.
  \item[LoR] Letter of Recommendation. A letter written by a professor or supervisor that assesses your qualifications and potential for graduate study (\autoref{sec:lor}).
  \item[SOP] Statement of Purpose. A document written by yourself to explains your research interests, background, and reasons for applying to a PhD program (\autoref{sec:sop}).
